Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-20 paper/paper.tex
checking before meeting

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic. First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse. Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

@@

@@

\begin{theorem}[\agdalogo]
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to $\mathsf{O}~0~\true$.
\end{theorem}
\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

@@

%% \subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}. Henglein and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of regular grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.

>>>>

In addition to these type-theoretic principles, we need two additional
axioms that do not generally hold in systems based on linear
logic.

\paragraph{Distributivity}
First, we need that additive conjunction \emph{distributes}
over additive disjunction --- e.g., in the finitary case that $0 \& A
\cong 0$ and $(A + B) \& C \cong (A \& C) + (B \& C)$. In its most
general form, the axiom says that the definable function
${\bigoplus\limits_{f : \prod_{x:X}
    Y(x)}\bigamp\limits_{x:X}A\,x\,(f\,x) \lto
  \bigamp\limits_{x:X}\bigoplus\limits_{y:Y(x)}A\,x\,y}$ has an
inverse.

\steven{Maybe this detail should go closer to the actual example
We use distributivity to encode the lookahead of the LL(1) parser in
\cref{thm:exp-parser}. At parse time, an LL(1) automaton peeks at the next
character of a string and uses the peeked character to inform its next
transition. Our automaton from \cref{fig:binop-inductive} chooses which state to
transition to dependent on if the next character is $\stringquote{)}$.

TODO

the proof obligation is

$\Char \otimes \LinPiTy{s}{state}{Trace s true}$ which is iso to
$(\literal{)} \otimes \top \oplus NotStatsWithRP) \otimes
\LinPiTy{s}{state}{Trace s true}$.

However, without distributivity we get stuck and can go further to actually use
$lookaheadRP$ or $lookaheadNot$.

}

\paragraph{Disjointness of Constructors}
Second, we need that the different constructors of
$\bigoplus$ are disjoint. We add this by adding the axiom that for any
$A : X \to L$ and $x \neq x' : X$ there is a function
$\uparrow(\equalizer{b}{\sigma\,x \circ \pi_1}{\sigma\,x'\circ \pi_2}
\lto 0)$ where $b: A(x) \& A(x')$, i.e., the grammar of pairs of an
$a:A(x)$ and an $a': A(x')$ such that $\sigma\,x\,a = \sigma\,x'\,a'$
is empty.

@@

\steven{To define $\DtoN$, we take the smallest trace through $\N$ subject to an
  ordering on its states and transitions, where the ordering is supplied
  non-linearly. In essence, the given ordering specifies a disambiguation strategy.}

@@

\begin{theorem}[\agdalogo]
  \label{thm:exp-parser}
  We construct a parser for $\mathsf{Exp}$ by showing it is weakly
  equivalent to $\mathsf{O}~0~\true$.
\end{theorem}
\begin{figure}
\begin{lstlisting}
data Exp : L where
  done : (*@$\uparrow$@*)(Atom (*@$\lto$@*) Exp)
  add : (*@$\uparrow$@*)(Atom (*@$\lto$@*) '+' (*@$\lto$@*) Exp (*@$\lto$@*) Exp)
data Atom : L where
  num : (*@$\uparrow$@*)(NUM (*@$\lto$@*) Atom)
  parens : (*@$\uparrow$@*)('(' (*@$\lto$@*) Exp (*@$\lto$@*) ')' (*@$\lto$@*) Atom)

@@

%% \subsubsection{Typed Grammars}
The usage of a simple type system to reason about regular expressions up to weak
equivalence was
introduced by Frisch and Cardelli \cite{frischCardelli}. \steven{Frisch and
  Caredilli also define a linear time greedy parser for regular expressions. We
  have not yet tried to verify a greedy algorithm, however the approach used in
  \cref{thm:determinization} would apply to their algorithm.
  \cref{thm:determinization} requires a choice function for traces through an
  NFA (i.e. a disambiguation strategy) which may be instantiated to encode greediness. } Henglein and Nielsen
\cite{henglein_regular_2011} later adapt this approach by providing a type
system for regular expressions whose semantics were in one-to-one correspondence
with the set of parse trees.
Elliott \cite{elliottSymbolicAutomaticDifferentiation2021} gives a similar
semantic interpretation of regular grammars as type-level predicates on strings. The type system of \theoryabbv extends these semantic interpretations of to a broader class of grammars and also gives a formal syntax and equational theory for the parse transformers that these prior works lack.
